(Coursera) Practical Machine Learning Project Write-up
==============================================================================

<big>
**Ivan Ngeow**  
**21 Jun 2014**
</big>

how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did

exploratory analysis
feature selection: NZV, removing NA columns, are the outcome classes balanced?
data preprocessing, partitioning
model training, performance on CV
model selection and fine-tuning
performance on validation set: can't ROC because 5 classes

Trade-off in training an accurate model due to time, CPU resources, memory.

```{r setup, include=FALSE}
opts_knit$set(progress=FALSE, verbose=FALSE, width=90)
opts_chunk$set(message=FALSE, tidy=TRUE, highlight=TRUE, fig.align="center")
```

```{r prepareData}
    a <- read.csv('pml-training.csv', stringsAsFactors=F)
    a$classe <- as.factor(a$classe)

    ## remove columns with NA
    na.counts <- apply(a, 2, function(x) length(which(is.na(x))))

    ## remove nzv columns
    nzv.cols <- nearZeroVar(a)#,saveMetrics=TRUE)

    exclude.cols <- union(which(na.counts!=0), c(1:7,nzv.cols))
    a <- a[, -exclude.cols ]

    ## partition
    inTrain <- createDataPartition(y=a$classe, p=0.85, list=T)[[1]]
    
    training <- a[inTrain,]
    testing <- a[-inTrain,]
```

A summary of Markdown syntax is available when editing R Markdown documents
under RStudio, or on this [Markdown Cheatsheet](http://goo.gl/6ThPR7).

To compile this document, you can use the following commands in R:

    library(knitr)
    library(markdown)
    knit("template.Rmd", quiet=TRUE)
    markdownToHTML("template.md", "template.html", option=c("highlight_code", "toc"), title="Simple R Markdown template")
    browseURL("template.html")

```{r mtry, fig.width=6, fig.height=6}
tc <- trainControl(method = "cv", number = 3, returnResamp='all')
small <- training[sample(1:nrow(training), 1000),]
plot(train(small[, pred.cols], small$classe, method='rf', ntree=2000, tuneGrid=data.frame(mtry=2:13), trControl=tc))
```

```{r train}
## use entire training set to train
tc <- trainControl(method = "cv", number = 3, returnResamp='all')
fit <- train(training[, pred.cols], training$classe, method='rf', ntree=2000, tuneGrid=data.frame(mtry=8:10), trControl=tc)

## now test on the hold-out set to obtain out-of-sample metrics
prediction <- predict(fit, newdata=testing)
print(confusionMatrix(prediction, testing$classe), digits=4)
```